strucplots---Visualizing higher-dimensional Contingency Tables in \pkg{vcd}

### 2005-08-24 trunk/vcd/inst/doc/strucplot.Rnw
New shading vignette added.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
viewport names changed to "obj:dim1=level1,...,dimn=leveln" syntax.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@444 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine=R,eps=FALSE,prefix.string=pics/strucplot}
\setkeys{Gin}{width=0.8\textwidth}

@@

<<structable,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
structable(Titanic)
@

Let us
first visualize the contingency table by the means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits (see Figure \ref{fig:observed}):
\begin{Sinput}
> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
\end{Sinput}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Observed,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, main = TRUE)
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent  It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):
\begin{Sinput}
> mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Expected,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to directly compare observed and expected values,
we can use a sieve plot (see Figure \ref{fig:sieve}):
\begin{Sinput}
> sieve(Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<sieve,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(Titanic)
@
\caption{Sieve plot for the Titanic data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent The next step is to directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):
\begin{Sinput}
> assoc(Titanic, compress = FALSE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Residuals,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(Titanic, compress = FALSE)
@
\caption{Association plot for the Titanic data.}
\label{fig:residuals}
\end{center}
\end{figure}

@@

So far, we have visualized the full table. Parts of the data can be
selected either using \code{margin.table()} on the input data, or the
formula interfaces. The following example compares the age
distribution among classes, given gender, and also shows how to 
use grid's viewport framework to put two plots besides 
each other:
\begin{Sinput}
> pushViewport(viewport(layout = grid.layout(ncol = 2)))

> pushViewport(viewport(layout.pos.col = 1))
> mosaic(~ Class + Age, data = Titanic[,"Male",,], 
+        margin = c(bottom = 0), sub = "Male", newpage = FALSE)
> popViewport()

> pushViewport(viewport(layout.pos.col = 2))
> mosaic(~ Class + Age, data = Titanic[,"Female",,], 
+        margin = c(bottom = 0), sub = "Female", newpage = FALSE)
> popViewport(2)
\end{Sinput}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Variables,fig=TRUE,echo=FALSE,height=6,width=16,results=hide>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))

pushViewport(viewport(layout.pos.col = 1))
mosaic(~ Class + Age, data = Titanic[,"Male",,], 
       margin = c(bottom = 0), sub = "Male", newpage = FALSE,
       labeling_args = list(gp_labels = gpar(fontsize = 24), 
                            gp_varnames = gpar(fontsize = 24)),
       title_gp = gpar(fontsize = 35))
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(~ Class + Age, data = Titanic[,"Female",,], 
       margin = c(bottom = 0), sub = "Female", newpage = FALSE,
       labeling_args = list(gp_labels = gpar(fontsize = 24), 
                            gp_varnames = gpar(fontsize = 24)),
       title_gp = gpar(fontsize = 35))
popViewport(2)
@
\caption{Age distribution among classes, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}

@@

\begin{Sinput}
> mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]

\begin{center}
<<Conditioning,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}

@@

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one). Sensible recycling rules apply. 
For compatibility with \code{mosaicplot()}, the \code{mosaic()} function
also allows the use of a \code{"direction"} argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):
\begin{Sinput}
> mosaic(Titanic, direction = "h")
> mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<split1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, direction = "h")
@
\caption{Mosaic plot for the Titanic data---alternative splitting (1).}
\label{fig:split1}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<split2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
@
\caption{Mosaic plot for the Titanic data---alternative splitting (2).}
\label{fig:split2}
\end{center}
\end{figure}

@@

The last plot pretty much resembles a doubledecker plot, except for
the spacing. Since spacing is also modular in the strucplot framework,
the \code{doubledecker()} function was straightforward to implement:
it really is just a wrapper for \code{mosaic()}, setting the right
defaults. Figure \ref{fig:doubledecker} shows a doubledecker plot of
the Titanic data, explaining the probability of Survival by Age, given
Sex, given Class:
\begin{Sinput}
> doubledecker(Titanic)
> ## equivalent to:
> doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<doubledecker,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(Titanic)
@
\caption{Doubledecker plot for the Titanic data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

@@

\begin{Sinput}
> pairs(Titanic, upper_panel = pairs_assoc)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
pairs(Titanic, upper_panel = pairs_assoc)
@
\caption{Pairs plot for the Titanic data.}
\label{fig:pairs}
\end{center}
\end{figure}

@@

\begin{Sinput}
> cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
\end{Sinput}

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplot,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
@
\caption{cotabplot for the Titanic data.}
\label{fig:cotabplot}
\end{center}
\end{figure}

@@

All strucplot components (including the panel functions) 
are parameterized through the use of generating functions, that is,
functions with all parameters in their signature returning a function
to be called by \code{strucplot()}. This returned function has access
to all parameters passed to the generating function thanks to 
lexical scope. The idea is to use module-specific `code
factories' that generate parameterized workhorses rather than
struggling with monolithic functions with endless parameter lists. As
a simple example, consider \code{struc\_mosaic()} and \code{struc\_assoc},
the generating functions for the mosaic and the association plots:

@@

\begin{Sinput}
> strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
> strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
> strucplot(Titanic, panel = myworkhorse)
\end{Sinput}

@@

\begin{Sinput}
> mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
> current.vpTree() ## output not shown

> seekViewport("cell..Sex.Male")
> grid.rect(gp = gpar(col = "red", lwd = 4))

> seekViewport("cell..Sex.Male..Class.Crew")
> grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
\end{Sinput}

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<afterplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)

seekViewport("cell..Sex.Male")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell..Sex.Male..Class.Crew")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@
\caption{Adding elements to a mosaic plot after drawing.}
\label{fig:afterplot}
\end{center}
\end{figure}

@@

\begin{Sinput}
> assoc(Titanic)
> getNames() ## output not shown
> grid.edit("rect..Class.1st..Age.Adult..Sex.Female..Survived.Yes", 
            gp = gpar(fill = "red"))
\end{Sinput}

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<changeplot,fig=TRUE,echo=FALSE,height=6,width=6,results=hide>>=
assoc(Titanic)
x = Titanic
x[] = NA
x["1st", "Female", "Adult", "Yes"] = "Red"
assoc(Titanic, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}

>>>>

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}

@@

<<structable,results=verbatim>>=
structable(Titanic)
@

Let us first visualize the contingency table by the means of a mosaic plot
\citep{vcd:Hartigan+Kleiner:1984} which is basically 
an area-proportional visualization of (typically observed) frequencies, composed
of tiles (corresponding to the cells) created by recursive
vertical and horizontal splits of a square. Thus, the area of each tile
is proportional to the corresponding cell entry \emph{given} the
dimensions of previous splits (see Figure \ref{fig:observed}):

<<Observed>>=
mosaic(Titanic, main = TRUE)
@

@@

\begin{figure}[h]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the Titanic data.}
\label{fig:observed}
\end{center}
\end{figure}

\noindent  It is also possible to visualize the expected values instead of the
observed values (see Figure \ref{fig:expected}):

<<Expected>>=
mosaic(Titanic, type = "expected", main = TRUE, sub = "Expected values")
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the Titanic data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

\noindent In order to directly compare observed and expected values,
we can use a sieve plot (see Figure \ref{fig:sieve}):

<<sieve>>=
sieve(Titanic)
@

\begin{figure}[h]
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE>>=
<<sieve>>
@
\caption{Sieve plot for the Titanic data visualizing simultaneously
  observed and expected values.}
\label{fig:sieve}
\end{center}
\end{figure}

\noindent The next step is to directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are preferably visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \code{assocplot()}, \pkg{vcd}'s \code{assoc()}
function scales to more than two variables (see Figure \ref{fig:residuals}):

<<Residuals>>=
assoc(Titanic, compress = FALSE)
@

\begin{figure}[h]
\begin{center}
<<Residualsfig,fig=TRUE,echo=FALSE>>=
<<Residuals>>
@
\caption{Association plot for the Titanic data.}
\label{fig:residuals}
\end{center}
\end{figure}

@@

So far, we have visualized the full table. Parts of the data can be
selected either using \code{margin.table()} on the input data, or the
formula interfaces. The following example compares the age
distribution among classes, given gender, and also shows how to 
use grid's viewport framework to put two plots besides 
each other:

<<Variables>>=
pushViewport(viewport(layout = grid.layout(ncol = 2)))

pushViewport(viewport(layout.pos.col = 1))
mosaic(~ Class + Age, data = Titanic[,"Male",,], 
       margin = c(bottom = 0, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(~ Class + Age, data = Titanic[,"Female",,], 
       margin = c(bottom = 0, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
<<Variables>>
@
\caption{Age distribution among classes, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

<<Conditioning>>=
mosaic(~ Class + Age | Sex, data = Titanic, split = TRUE)
@

\begin{figure}[h]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}

@@

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument (default: alternating splits starting
with a vertical one). Sensible recycling rules apply. 
For compatibility with \code{mosaicplot()}, the \code{mosaic()} function
also allows the use of a \code{"direction"} argument taking a vector of \code{"h"} and
\code{"v"} characters (see Figures \ref{fig:split1} and \ref{fig:split2}):

<<split1>>=
mosaic(Titanic, direction = "h")
@

<<split2>>=
mosaic(Titanic, split_vertical = c(TRUE, TRUE, TRUE, FALSE),
                labeling_args = list(abbreviate = c(Sex = TRUE, Age = TRUE)))
@

\begin{figure}[h]
\begin{center}
<<split1fig,fig=TRUE,echo=FALSE>>=
<<split1>>
@
\caption{Mosaic plot for the Titanic data---alternative splitting (1).}
\label{fig:split1}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
<<split2fig,fig=TRUE,echo=FALSE>>=
<<split2>>
@
\caption{Mosaic plot for the Titanic data---alternative splitting (2).}
\label{fig:split2}
\end{center}
\end{figure}

@@

The last plot pretty much resembles a doubledecker plot, except for
the spacing. Since spacing is also modular in the strucplot framework,
the \code{doubledecker()} function was straightforward to implement:
it really is just a wrapper for \code{mosaic()}, setting the right
defaults. Figure \ref{fig:doubledecker} shows a doubledecker plot of
the Titanic data, explaining the probability of Survival by Age, given
Sex, given Class:

<<doubledecker1>>=
doubledecker(Titanic)
@

\noindent equivalent to:

<<doubledecker2>>=
doubledecker(Survived ~ Class + Sex + Age, data = Titanic)
@

\begin{figure}[h]
\begin{center}
<<doubledeckerfig,fig=TRUE,echo=FALSE>>=
<<doubledecker1>>
@
\caption{Doubledecker plot for the Titanic data.}
\label{fig:doubledecker}
\end{center}
\end{figure}

@@

<<pairs>>=
pairs(Titanic, upper_panel = pairs_assoc)
@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<pairsfig,fig=TRUE,echo=FALSE>>=
<<pairs>>
@
\caption{Pairs plot for the Titanic data.}
\label{fig:pairs}
\end{center}
\end{figure}

@@

<<cotabplot>>=
cotabplot(~ Class + Survived | Sex + Age, data = Titanic)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[h]
\begin{center}
<<cotabplotfig,fig=TRUE,echo=FALSE,height=6,width=6>>=
<<cotabplot>>
@
\caption{cotabplot for the Titanic data.}
\label{fig:cotabplot}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

All strucplot components (including the panel functions) 
are parameterized through the use of generating functions, that is,
functions with all parameters in their signature returning a function
to be called by \code{strucplot()}. This returned function has access
to all parameters passed to the generating function because in {\sf
  R}, nested functions are lexically scoped. The idea is to use module-specific `code
factories' that generate parameterized workhorses rather than
struggling with monolithic functions with endless parameter lists. As
a simple example, consider \code{struc\_mosaic()} and \code{struc\_assoc},
the generating functions for the mosaic and the association plots:

@@

<<panelfuns>>=
strucplot(Titanic, panel = struc_mosaic(zero_size = 1))
strucplot(Titanic, panel = struc_mosaic, panel_args = list(zero_size = 1))
@ 
<<panelfuns2,eval=FALSE>>=
strucplot(Titanic, panel = myworkhorse)
@

@@

<<viewportnames>>=
mosaic(~ Sex + Class, data = Titanic, pop = FALSE)
current.vpTree() ## output not shown

seekViewport("cell:Sex=Male")
grid.rect(gp = gpar(col = "red", lwd = 4))

seekViewport("cell:Sex=Male,Class=Crew")
grid.circle(r = 0.2, gp = gpar(fill = "cyan"))
@

\setkeys{Gin}{width=0.6\textwidth}
\begin{figure}[h]
\begin{center}
<<viewportnamesfig,fig=TRUE,echo=FALSE>>=
<<viewportnames>>
@
\caption{Adding elements to a mosaic plot after drawing.}
\label{fig:afterplot}
\end{center}
\end{figure}

@@

<<changeplot>>=
assoc(Titanic)
getNames() ## output not shown
grid.edit("rect:Class=1st,Age=Adult,Sex=Female,Survived=Yes", 
            gp = gpar(fill = "red"))
@

\begin{figure}[h]
\begin{center}
<<changeplotfig,fig=TRUE,echo=FALSE>>=
assoc(Titanic)
x = Titanic
x[] = NA
x["1st", "Female", "Adult", "Yes"] = "Red"
assoc(Titanic, gp = gpar(fill = x))
@
\caption{Changing graphical parameters of elements after drawing.}
\label{fig:changeplot}
\end{center}
\end{figure}
